---
title: "Moneyline Data Wrangling Part II"
author: "Greg Matthews and Mike Lopez"
date: "June 24, 2016"
output: html_document
---
  
```{r, message=FALSE}
require(mosaic)
```

Load the Results Data

### MLB

```{r}
mlb_results <- read.csv("~/Dropbox/competitiveness/Data/MLB_results.csv")
require(lubridate)
require(stringr)
mlb_res <- mlb_results %>%
  filter(Rk != "Rk") %>%
  # remove all the "road games"
  filter(X. != "@") %>%
  # capture doubleheaders
  mutate(dh = str_extract(str_extract(Date, "\\(.\\)"), "[0-9]+")) %>%
  mutate(dateString = paste(gsub("\\(.\\)", "", Date), year, sep = ",")) %>%
  mutate(gameDate = parse_date_time(dateString, orders = c("%b %d,%y"))) %>%
  mutate(home_team_id = Tm) %>%
  mutate(home_win = ifelse(as.numeric(R) > as.numeric(RA), TRUE, FALSE)) %>%
  mutate(sport = "mlb") %>%
  mutate(season = year) %>%
  mutate(playoffs = 0) %>%
  select(sport, season, gameDate, dh, home_team_id, home_win, playoffs)
```

Sanity check: should be $162 * 30 / 2 = 2430$ games per year

```{r}
mlb_res %>%
  group_by(season) %>%
  summarise(N = n(), numTeams = length(unique(home_team_id)), home_p = sum(home_win) / n()) %>%
  print.data.frame()
```


### NBA

```{r}
nba_results <- read.csv("~/Dropbox/competitiveness/Data/NBAdata.csv")
nba_res <- nba_results %>%
  mutate(dh = NA) %>%
  mutate(gameDate = parse_date_time(Date, orders = c("%b %d %Y"))) %>%
  mutate(home_team = Home.Neutral) %>%
  mutate(home_win = ifelse(as.numeric(PTS.1) > as.numeric(PTS), TRUE, FALSE)) %>%
  mutate(sport = "nba") %>%
  mutate(season = ifelse(month(gameDate) < 8, year(gameDate), year(gameDate) + 1)) %>%
  select(sport, season, gameDate, dh, home_team, home_win, playoffs)
# did we get all the dates?
nba_res %>%
  filter(is.na(gameDate))
```

### NFL

```{r}
nfl_results <- read.csv("~/Dropbox/competitiveness/Data/NFL_results.csv")
nfl_res <- nfl_results %>%
  na.omit() %>%
  mutate(dh = NA) %>%
  mutate(gameDay = parse_date_time(Date, orders = c("%b %d"))) %>%
  mutate(gameYear = ifelse(month(gameDay) < 6, season + 1, season)) %>%
  mutate(gameDate = parse_date_time(paste(Date, gameYear), orders = c("%b %d %y"))) %>%
  mutate(home_team = ifelse(Var.7 == "@", as.character(Loser.tie), as.character(Winner.tie))) %>%
  mutate(home_win = ifelse(Winner.tie == home_team, TRUE, FALSE)) %>%
  mutate(sport = "nfl") 

##Week of season
min.day <- nfl_res %>%
    group_by(season) %>%
    summarise(min.day = min(gameDate))

nfl_res <- nfl_res %>%
    left_join(min.day) %>%
    mutate(day = as.Date(gameDate) - as.Date(min.day), week = as.numeric(floor(day/7)+1), 
           playoffs = as.numeric(week > 17))
head(nfl_res)
  
nfl_res <- nfl_res %>%
  select(sport, season, gameDate, dh, home_team, home_win, playoffs)

### Looks good after 2004: 11 playoff games per year
table(nfl_res$playoffs, nfl_res$season)
```

### NHL

```{r}
nhl_results <- read.csv("~/Dropbox/competitiveness/Data/NHL_results.csv")
nhl_res <- nhl_results %>%
  mutate(dh = NA) %>%
  mutate(gameDate = ymd(Date)) %>%
  mutate(home_team = Home) %>%
  # what to do about ties?
  mutate(home_win = ifelse(G.1 >= G, TRUE, FALSE)) %>%
  mutate(sport = "nhl") %>%
  mutate(season = ifelse(month(gameDate) < 8, year(gameDate), year(gameDate) + 1)) %>%
  select(sport, season, gameDate, dh, home_team, home_win, playoffs)
# did we get all the dates?
nhl_res %>%
  filter(is.na(gameDate))
```

## Creating the Lookup

### MLB

Errors in Lahman package

```{r}
require(dplyr)
require(Lahman)
Teams %>%
  filter(teamID %in% c("CHA", "CHN") & yearID > 2010) %>%
  select(yearID, teamID, name)
```

```{r}
teamname_lkup <- Teams %>%
  filter(yearID >= min(year(mlb_res$gameDate))) %>%
  select(teamID, name) %>%
  unique()
# remove erroneous team pairs
teamname_lkup <- teamname_lkup[1:34,]
teamname_lkup <- bind_rows(teamname_lkup, data.frame(teamID = "TBD", name = "Tampa Bay Rays"))
teamname_lkup <- teamname_lkup %>%
  mutate(teamID = ifelse(teamID == "CHA", "CHW", as.character(teamID))) %>%
  mutate(teamID = ifelse(teamID == "NYA", "NYY", teamID)) %>%
  mutate(teamID = ifelse(teamID == "SLN", "STL", teamID)) %>%
  mutate(teamID = ifelse(teamID == "KCA", "KCR", teamID)) %>%
  mutate(teamID = ifelse(teamID == "CHN", "CHC", teamID)) %>%
  mutate(teamID = ifelse(teamID == "NYN", "NYM", teamID)) %>%
  mutate(teamID = ifelse(teamID == "FLO", "FLA", teamID)) %>%
  mutate(teamID = ifelse(teamID == "SFN", "SFG", teamID)) %>%
  mutate(teamID = ifelse(teamID == "LAN", "LAD", teamID)) %>%
  mutate(teamID = ifelse(teamID == "SDN", "SDP", teamID)) %>%
  mutate(teamID = ifelse(teamID == "TBA", "TBR", teamID)) %>%
  mutate(teamID = ifelse(teamID == "WAS", "WSN", teamID)) %>%
  mutate(name = ifelse(teamID == "FLA", "Miami Marlins", name)) %>%
  mutate(name = ifelse(teamID == "STL", "St Louis Cardinals", name)) %>%
  mutate(name = ifelse(teamID == "LAA", "Los Angeles Angels", name)) %>%
  mutate(name = ifelse(teamID == "TBR", "Tampa Bay Rays", name)) %>%
  rename(home_team = name) %>%
  unique()

mlb_res <- left_join(x = mlb_res, y = teamname_lkup, by = c("home_team_id" = "teamID"))
```

### NBA

In the money line data they were always the Brooklyn Nets and the Charlotte Hornets and the New Orleans Pelicans.

```{r}
nba_res %>%
  filter(grepl("Nets", home_team)) %>%
  select(home_team) %>%
  unique()
nba_res <- nba_res %>%
  mutate(home_team = ifelse(grepl("Nets", home_team), "Brooklyn Nets", as.character(home_team))) %>%
  mutate(home_team = ifelse(grepl("Bobcats", home_team), "Charlotte Hornets", as.character(home_team))) %>%
  mutate(home_team = ifelse(grepl("New Orleans.+Hornets", home_team), "New Orleans Pelicans", as.character(home_team)))

```

### NFL

### NHL


### Combine them all

```{r}
#results <- bind_rows(select(mlb_res, -home_team_id), nba_res, nfl_res, nhl_res)
results <- rbind(select(mlb_res, -home_team_id), nba_res, nfl_res, nhl_res)
```


## Load the Moneyline Data

```{r}
load("~/Dropbox/competitiveness/Data/mlineGM.rda")

# Check dates for mline data
mline$month <- month(mline$gameDate)
mline %>% 
  group_by(sport, season) %>% 
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>%
  print.data.frame()

```

Account for differences in team names between the two data sets


Last minute addendums:
- Seattle Supersonics should be merged with OKC Thunder (NBA). Supersonics spelt two ways
- Atlanta Thrashers should be merged with Winnipeg Jets
- Coyotes (NHL) are Arizona on the road and Phoenix at home
- Ducks (NHL) also come in two forms

```{r}
Seattle <- c("Seattle Supersonics", "Seattle SuperSonics")
mline[mline$visitor_team %in% Seattle,]$visitor_team <- "Oklahoma City Thunder"
mline[mline$home_team %in% Seattle,]$home_team <- "Oklahoma City Thunder"

mline[mline$visitor_team=="Atlanta Thrashers",]$visitor_team <- "Winnipeg Jets"
mline[mline$home_team=="Atlanta Thrashers",]$home_team <- "Winnipeg Jets"

#mline[mline$visitor_team=="Mighty Ducks of Anaheim",]$visitor_team <- "Anaheim Ducks"
#mline[mline$home_team=="Mighty Ducks of Anaheim",]$home_team <- "Anaheim Ducks"

#mline[mline$visitor_team=="Phoenix Coyotes",]$visitor_team <- "Arizona Coyotes"
#mline[mline$home_team=="Phoenix Coyotes",]$home_team <- "Arizona Coyotes"


Seattle <- c("Seattle Supersonics", "Seattle SuperSonics")

results[results$home_team %in% Seattle,]$home_team <- "Oklahoma City Thunder"
results[results$home_team=="Atlanta Thrashers",]$home_team <- "Winnipeg Jets"
results[results$home_team=="Mighty Ducks of Anaheim",]$home_team <- "Anaheim Ducks"
results[results$home_team=="Phoenix Coyotes",]$home_team <- "Arizona Coyotes"

```

```{r}
bigfour <- right_join(x = mline, y = results, by = c("sport", "season", "gameDate", "home_team"))

teams.1 <- unique(filter(bigfour, season > 2004)$home_team)
teams.2 <- unique(filter(bigfour, season > 2004)$visitor_team)
teams.2 %in% teams.1 ##The NA is missing
```

How many games do we have?

```{r}
bigfour %>%
  group_by(sport, season) %>%
  summarise(N_results = n(), numTeams = length(unique(setdiff(union(visitor_team, home_team), NA))), 
            G = sum(!is.na(home_win)), W = sum(home_win, na.rm = TRUE), L = sum(!home_win, na.rm = TRUE),
            home_wp = sum(home_win, na.rm = TRUE) / sum(!is.na(home_win)),
            N_mline = sum(!is.na(prob_home)), 
            mean_home_p = mean(prob_home, na.rm = TRUE)) %>%
  mutate(coverage = N_mline / N_results) 
```

```{r}
bigfour %>%
  filter(year(gameDate) == 2012 & sport == "nba") %>%
  group_by(home_team) %>%
  summarise(N = n()) %>%
  arrange((desc(N))) 

nba_res %>% 
  filter(sport == "nba") %>%
  filter(grepl("Bobcats|Hornets|New Orleans", home_team)) %>%
  group_by(year(gameDate), home_team) %>%
  summarise(N = n()) 

mline %>% 
  filter(sport == "nba") %>%
  filter(grepl("Bobcats|Hornets|New Orleans", home_team)) %>%
  group_by(year(gameDate), home_team) %>%
  summarise(N = n()) 

bigfour %>%
  filter(sport == "nhl" & season == 2006) 
mline %>%
  filter(sport == "nhl" & gameDate == "2005-10-05")
```

We have nearly all of them from 2005-2014. 

```{r}
bigfour %>%
  filter(season %in% 2005:2014) %>%
  group_by(sport, season) %>%
  summarise(N_results = n(), 
            numTeams = length(unique(setdiff(union(visitor_team, home_team), NA))), 
            home_wp = sum(home_win, na.rm = TRUE) / sum(!is.na(home_win)),
            N_mline = sum(!is.na(prob_home)), 
            mean_home_p = mean(prob_home, na.rm = TRUE)) %>%
  mutate(coverage = N_mline / N_results) %>%
  print.data.frame()
```

```{r}
bigfour <- bigfour %>%
  filter(season %in% 2005:2014) %>%
  # four hockey games that were postponed
  filter(!is.na(home_win))
summary(bigfour)
```

We're missing only a tiny fraction of the games:

```{r}
bigfour %>%
summarize(coverage = sum(!is.na(p_home)) / n())

bigfour <- filter(bigfour, !is.na(vig))

bigfour %>%
summarize(coverage = sum(!is.na(p_home)) / n())
```


Check maximum dates within each sport (look for July NBA and NHL games, look for weird preseason dates)

```{r}
bigfour %>% 
group_by(sport, season) %>% 
summarise(min.date = min(gameDate), max.date = max(gameDate)) %>% 
print.data.frame()
```



```{r}
table(bigfour$home_team, bigfour$season)
table(bigfour$visitor_team, bigfour$season)
bigfour <- select(bigfour, -dh.x, -dh.y)
save(bigfour, file = "~/Dropbox/competitiveness/Data/bigfourGM.rda", compress = "xz")
```
