---
title: "Moneyline Data Wrangling"
author: "Mike Lopez"
date: "October 4, 2016"
output: html_document
---

```{r, message=FALSE}
source("../config.R")
library(lubridate)
nhl <- read_csv(file.path(data_raw, "NHL.csv"))
nba <- read_csv(file.path(data_raw, "NBA.csv"))
nfl <- read_csv(file.path(data_raw, "NFL.csv"))
mlb <- read_csv(file.path(data_raw, "MLB.csv"))

vars <- c("Date", "Visitor Team", "Home Team", "Visitor ML", "Home ML", "Visitor Score", "Home Score")

nhl <- nhl[,vars]
nba <- nba[,vars]
nfl <- nfl[,vars]
mlb <- mlb[,vars]
```


Combine them

```{r, message=FALSE}
ds <- bind_rows(mlb, nba, nfl, nhl)
ds <- mutate(ds, sport = c(rep("mlb", nrow(mlb)), 
                           rep("nba", nrow(nba)), 
                           rep("nfl", nrow(nfl)), 
                           rep("nhl", nrow(nhl))))
```

Remove any duplicate lines

```{r}
ds <- unique(ds)
```

Clean up dates

```{r, message=FALSE}
ds <- ds %>%
  mutate(gameDate = mdy_hm(Date), gameTime = mdy_hm(Date)) %>%
# discard games with no times? For now???
  filter(!is.na(gameTime))
```

Make the columns easier to use

```{r}
ds <- ds %>%
  rename(ml_vis = `Visitor ML`, ml_home = `Home ML`, 
         visitor_team = `Visitor Team`, home_team = `Home Team`, 
         visitor_score = `Visitor Score`, home_score = `Home Score`)
```

Define the seasons

```{r}
# start with just using the year
ds <- ds %>%
  mutate(season = year(gameDate))
# but for NBA and NHL, have to adjust
ds <- ds %>%
  mutate(season = ifelse(sport %in% c("nba", "nhl") & month(gameDate) > 7, 
                         year(gameDate) + 1, season))
# NFL too
ds <- ds %>%
  mutate(season = ifelse(sport == "nfl" & month(gameDate) < 4, 
                season - 1, season))
```

Strip out all games where we don't have the line

```{r}
ds <- filter(ds, !is.na(ml_home))
```

Find all the real teams

```{r}
teams <- ds %>%
  group_by(home_team) %>%
  summarise(N = n()) %>%
  filter(N > 100) %>%
  arrange(desc(N)) %>%
  filter(home_team!="Home Goals")

teams1 <- ds %>%
  group_by(visitor_team) %>%
  summarise(N = n()) %>%
  filter(N > 100) %>%
  arrange(desc(N)) 
```

Restrict to only to those teams

```{r}
ds <- semi_join(ds, teams, by = c("home_team"))
ds <- semi_join(ds, teams1, by = c("visitor_team"))
```


Still have multiple lines for some games!

```{r}
dupes <- ds %>%
  group_by(home_team, gameTime) %>%
  summarise(N = n()) %>%
  filter(N > 1)
dupes
```

Arbitrarily remove one of them

```{r}
dup.idx <- ds %>%
  semi_join(dupes, by = c("gameTime", "home_team"))
# just take the later ones
dup.bad <- dup.idx[rep(c(TRUE, FALSE), nrow(dup.idx) / 2),]
ds <- ds %>%
  anti_join(dup.bad, by = c("gameTime", "home_team"))
```


What to do about doubleheaders? 

```{r}
filter(ds, sport == "mlb" & season == 2013 & 
         home_team == "Colorado Rockies" & 
         month(gameDate) == 4 & 
         day(gameDate) == 16) %>% 
  print.data.frame()
dh <- ds %>%
  group_by(gameDate, home_team) %>%
  summarise(numGames = n()) %>%
  filter(numGames > 1)
dh.idx <- dh %>%
  semi_join(x = ds, by = c("gameDate", "home_team")) %>%
  mutate(dh = rep(c(1,2), nrow(dh)))

ds <- ds %>%
  left_join(select(dh.idx, gameTime, home_team, dh), by = c("gameTime", "home_team"))

# no tripleheaders
ds <- filter(ds, is.na(dh) == TRUE | dh < 2)

ds %>%
  group_by(gameDate, home_team) %>%
  summarise(numGames = n()) %>%
  filter(numGames > 2)
```

Note that there are some MLB regular season games in March

```{r}
ds %>% mutate(month = month(gameDate)) %>%
  group_by(month, sport) %>%
  summarise(N = n()) %>%
  print.data.frame()

ds <- ds %>%
  filter(!(sport == "mlb" & month(gameDate) < 3)) %>%
  filter(!(sport == "nba" & month(gameDate) < 10 & month(gameDate) > 6)) %>%
  filter(!(sport == "nfl" & month(gameDate) > 2 & month(gameDate) < 9)) %>%
  filter(!(sport == "nhl" & month(gameDate) < 9 & month(gameDate) > 6)) %>%
  filter(season > 2005, season < 2017)
```

```{r}
ds %>%
  group_by(sport, season) %>%
  summarise(N = n(), numTeams = length(unique(union(visitor_team, home_team)))) %>%
  print.data.frame()
```


Compute implied probabilities

```{r}
ds <- ds %>%
  mutate(prob_vis = ifelse(ml_vis > 0, 
                           100 / (100 + ml_vis), 
                           abs(ml_vis) / (100 + abs(ml_vis)))) %>%
  mutate(prob_home = ifelse(ml_home > 0, 
                            100 / (100 + ml_home), 
                            abs(ml_home) / (100 + abs(ml_home)))) %>%
  mutate(vig = (prob_vis + prob_home) - 1) %>%
  mutate(p_vis = prob_vis / (prob_vis + prob_home)) %>%
  mutate(p_home = prob_home / (prob_vis + prob_home))
```

# double-check for bad data

```{r}
test <- filter(ds, ml_vis > 0 & p_vis > 0.5)
test
test$ml_vis

test <- filter(ds, vig > 0.1)
test
test$ml_vis
test$ml_home
```

Get rid of these problematic money lines 

```{r}
ds <- filter(ds, !ml_vis==56, !ml_home ==65)
```


Save the resulting data frame

```{r mline}
mline <- ds
#save(mline, file = "~/Dropbox/Competitiveness/SportsInsights/mline_final.rda", compress = "xz")
```


At this point, we have a near-final set of money line data, now we just have to figure out which games are RS. 

We'll do this by using our results files, as well as some look-ups of recent seasons. 


```{r}
mlb_results <- read_csv(file.path(root, "data", "MLB_results.csv"))
mlb.new <- data.frame(season = c(2015, 2016), 
                      min.date = parse_date_time(c("2015-04-05", "2016-04-03"), "ymd"), 
                      max.date = parse_date_time(c("2015-10-04", "2016-10-02"), "ymd"))
mlb_dates <- mlb_results %>%
  mutate(dateString = paste(gsub("\\(.\\)", "", Date), year, sep = ","),
         gameDate = parse_date_time(dateString, orders = c("%b %d,%y")),
         season = year) %>%
  group_by(season) %>%
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>%
  bind_rows(mlb.new) %>%
  mutate(sport = "mlb") 
mlb_dates[mlb_dates$season == 2014,]$min.date <- 
  parse_date_time("2014-03-31", "ymd")
mlb_dates[mlb_dates$season == 2012,]$min.date <- 
  parse_date_time("2012-04-04", "ymd")


nhl_results <- read_csv(file.path(root, "data", "NHL_results.csv"))
nhl.new <- data.frame(season = c(2016), 
                      min.date = parse_date_time(c("2015-10-07"), "ymd"), 
                      max.date = parse_date_time(c("2016-04-10"), "ymd"))
nhl_dates <- nhl_results[,c(2, 11)] %>%
  filter(playoffs == 0) %>%
  mutate(gameDate = parse_date_time(Date, "ymd"),
         season = ifelse(month(gameDate) < 8, year(gameDate), year(gameDate) + 1))  %>%
  group_by(season) %>%
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>%
  bind_rows(nhl.new) %>%
  mutate(sport = "nhl")


nba_results <- read_csv(file.path(root, "data", "NBA_results.csv"))
nba.new <- data.frame(season = c(2015, 2016), 
                      min.date = parse_date_time(c("2014-10-28", "2015-10-27"), "ymd"), 
                      max.date = parse_date_time(c("2015-04-15", "2016-04-16"), "ymd"))

nba_dates <- nba_results[,c(2,11)] %>%
  filter(playoffs == 0) %>%
  mutate(gameDate = parse_date_time(Date, orders = c("%b %d %Y")),
         season = ifelse(month(gameDate) < 8, year(gameDate), year(gameDate) + 1)) %>%
  group_by(season) %>%
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>%
  bind_rows(nba.new) %>%
  mutate(sport = "nba")


nfl_results <- read_csv(file.path(root, "data", "NFL_results.csv"))
nfl.new <- data.frame(season = 2015, 
                      min.date = parse_date_time("2015-09-10", "ymd"), 
                      max.date = parse_date_time("2016-01-03", "ymd"))
nfl_dates <- nfl_results %>%
  na.omit() %>%
  filter(Week!="Champ", Week!="ConfChamp", Week!="Division", Week!="SuperBowl", Week!="WildCard") %>%
  mutate(gameDay = parse_date_time(Date, orders = c("%b %d")),
         gameYear = ifelse(month(gameDay) < 6, season + 1, season),
         gameDate = parse_date_time(paste(Date, gameYear), orders = c("%b %d %y"))) %>%
  group_by(season) %>%
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>%
  bind_rows(nfl.new) %>%
  mutate(sport = "nfl") 
  
all.dates <- bind_rows(nfl_dates, nhl_dates, mlb_dates, nba_dates)
```

Now merge

```{r}
mline.dates <- left_join(mline, all.dates)
mline.dates <- mline.dates %>%
  filter(gameDate %within% lubridate::interval(min.date, max.date + days(1)))
```

MLB plays 2430 games in a year, NHL & NBA 1230, NFL 240

```{r}
mline.dates %>%
  group_by(sport, season) %>%
  summarise(N = n()) %>% print.data.frame()

mline.dates %>%
  group_by(home_team, season) %>%
  summarise(N = n()) %>% print.data.frame()
```

Two franchise switches

```{r}
mline.dates[mline.dates$visitor_team == "Atlanta Thrashers",]$visitor_team <- "Winnipeg Jets"
mline.dates[mline.dates$home_team == "Atlanta Thrashers",]$home_team <- "Winnipeg Jets"

mline.dates[mline.dates$visitor_team == "Seattle Supersonics",]$visitor_team <- "Oklahoma City Thunder"
mline.dates[mline.dates$home_team == "Seattle Supersonics",]$home_team <- "Oklahoma City Thunder"
```

Looking at data

```{r}
bigfour <- mline.dates
bigfour %>%
  group_by(sport, season) %>%
  mutate(home_win = home_score > visitor_score) %>%
  summarise(N_results = n(), numTeams = length(unique(setdiff(union(visitor_team, home_team), NA))), 
            G = sum(!is.na(home_win)), W = sum(home_win, na.rm = TRUE), L = sum(!home_win, na.rm = TRUE),
            home_wp = sum(home_win, na.rm = TRUE) / sum(!is.na(home_win)),
            N_mline = sum(!is.na(prob_home)), 
            mean_home_p = mean(prob_home, na.rm = TRUE)) %>%
  mutate(coverage = N_mline / N_results) %>%
  print.data.frame()


bigfour %>% 
  group_by(sport, season) %>% 
  summarise(min.date = min(gameDate), max.date = max(gameDate)) %>% 
  print.data.frame()
```

Save it!

```{r}
bigfour <- bigfour %>%
  mutate(home_win = home_score > visitor_score)
save(bigfour, file = file.path(data_raw, "bigfour.rda"), compress = "xz")
```

Create a public version of the data set

```{r}
bigfour_public <- bigfour %>%
  select(gameDate, visitor_team, home_team, visitor_score, home_score, p_home)
save(bigfour_public, file = file.path(root, "data", "bigfour_public.rda"), compress = "xz")
```
