---
title: "JASA ACS Reproducibility Initiative - Author Contributions Checklist Form"
date: "January 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
source("../config.R")
knitr::opts_chunk$set(echo = TRUE)
```


## Data

### Abstract (Mandatory)
Our raw data was purchased from Sports Insights. Each observation is a game in one of the four major North American sports leagues (MLB, NBA, NFL, NHL) between 2005-2016, and contains the game date, the two participating teams, the final score, and the betting lines taken by Sports Insights before the game. 

We wrangled these raw data---stripping out duplicate entries, cleaning up team names, etc.---into a single master R `data.frame` called `bigfour.rda`. These data contain the betting lines and therefore cannot be shared. We have placed this file in a private directory denoted `data_raw`.

```{r}
load(file.path(data_raw, "bigfour.rda"))
dim(bigfour)
```

There are `r format(nrow(bigfour), scientific = FALSE, big.mark = ",")` observations and `r ncol(bigfour)` variables in `bigfour`.

A derivative public data frame with the same number of observations but many fewer variables is called `bigfour_public` and is available in our GitHub repository. 

```{r}
load(file.path("..", "data", "bigfour_public.rda"))
glimpse(bigfour_public)
```

This data frame contains only the game result and the implied probability of the home team winning, based on the betting line. 

### Availability (Mandatory)
The raw data purchased from Sports Insights is proprietary and cannot be shared publicly. However, we have created a summarized version of these data that is publicly shared. 

Description (Mandatory if data available)
Permissions (demonstrate that author has legitimate access to data)
Licensing information
Link to data (e.g., dataverse.org, datadryad.org; this need not be the actual link at time of submission but if not, it should indicate where the data will be deposited if the manuscript is accepted)
Data provenance, including identifier or link to original data if different than above
File format
Metadata (including data dictionary)
Version information

Optional Information (complete as necessary) 
Unique identifier / DOI

## Code
Abstract (Mandatory)
Short high level description

Description (Mandatory)
How delivered (R package, Shiny app, etc.)
Licensing information (default is MIT License)
Link to code/repository (e.g., github.com, bitbucket.org; this need not be the actual link at time of submission but if not, it should indicate where the code will be deposited if the manuscript is accepted)
Version information (e.g., for a Git repository, the number or branch+commit)

Optional Information (complete as necessary) 
Hardware requirements (e.g., operating system with version number, access to cluster, GPUs, etc.)
Supporting software requirements (e.g., libraries and dependencies, including version numbers)
Unique identifier/DOI 

## Instructions for Use
Reproducibility (Mandatory)
What is to be reproduced (e.g., "All tables and figure from paper", "Tables 1-4”, etc.)
How to reproduce analyses (e.g., workflow information, makefile, wrapper scripts)

Replication (Optional)
How to use software in other settings (or links to such information, e.g., R package vignettes, demos or other examples)

## Notes
Other relevant information, in particular how to access the data and code if not yet made publicly available.