---
title: "JASA ACS Reproducibility Initiative - Author Contributions Checklist Form"
date: "January 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
source("../config.R")
knitr::opts_chunk$set(echo = TRUE)
```


## Data

### Abstract (Mandatory)
Our raw data was purchased from Sports Insights. Each observation is a game in one of the four major North American sports leagues (MLB, NBA, NFL, NHL) between 2005-2016, and contains the game date, the two participating teams, the final score, and the betting lines taken by Sports Insights before the game. 

We wrangled these raw data---stripping out duplicate entries, cleaning up team names, etc.---into a single master R `data.frame` called `bigfour.rda`. These data contain the betting lines and therefore cannot be shared. We have placed this file in a private directory denoted `data_raw`.

```{r}
load(file.path(data_raw, "bigfour.rda"))
dim(bigfour)
```

There are `r format(nrow(bigfour), scientific = FALSE, big.mark = ",")` observations and `r ncol(bigfour)` variables in `bigfour`.

A derivative public data frame with the same number of observations but many fewer variables is called `bigfour_public` and is available in our GitHub repository. 

```{r}
load(file.path("..", "data", "bigfour_public.rda"))
glimpse(bigfour_public)
```

This data frame contains only the game result and the implied probability of the home team winning, based on the betting line. 

### Availability (Mandatory)

The raw data purchased from Sports Insights (i.e., `bigfour`) is proprietary and cannot be shared publicly. However, we have created a summarized version of these data (i.e., `bigfour_public`) that is publicly shared. 

### Description (Mandatory if data available)

- Permissions: Permission to share a version of the Sports Insights data was obtained via email during January, 2017
- Licensing information: Terms of use for Sports Insights data can be found at (https://www.sportsinsights.com/company/terms-of-use/)
- Link to data: (https://github.com/bigfour/competitiveness/data)
- Data provenance, including identifier or link to original data if different than above: The historical database of data was purchased at (https://www.sportsinsights.com/get-historical-betting-odds/)
- File format: `bigfour_public.rda` is an R data file
- Metadata (including data dictionary): Here's a snapshot of our variables:

Variable | Description
------------- | -------------
`gameDate` | Day of the game
`sport (q)` | League (MLB, NBA, NFL, or NHL)
`visitor_team (j)` | Visiting team 
`home_team (i)` | Home team 
`visitor_score` | Final score, visiting team
`home_score` | Final score, home team
`p_home` | $p_{q,i,j}$

where $p_{q,i,j}$ is the probability of team home team $i$ defeating away team $j$ in sport $q$. 

- Version information: Data purchased on October 6, 2016

## Code

### Abstract (Mandatory)
All of our code is hosted in a GitHub repository. 

### Description (Mandatory)

- How delivered: GitHub repository
- Licensing information: CC0
- Link to code/repository: (https://github.com/bigfour/competitiveness)
- Version information (e.g., for a Git repository, the number or branch+commit): **to be filled in just before submission**

### Optional Information (complete as necessary) 

- Supporting software requirements:
    - A recent version of R
    - [JAGS](http://mcmc-jags.sourceforge.net/)
    - CRAN packages: `tidyverse`, `rjags`, `knitr`, `lubridate`, `stringr`, `xtable`, `XML`, `RCurl`, `zoo`
    - GitHub packages: [`beanumber/teamcolors`](https://github.com/beanumber/teamcolors)
    - `pdflatex`

## Instructions for Use

### Reproducibility (Mandatory)

Please see the [README](https://github.com/bigfour/competitiveness/blob/master/README.md) for complete instructions. The first step in reproducing the analysis is to clone the GitHub repository. 

> Please note that none of these scripts will actually work without the original proprietary raw data. 

- What is to be reproduced:
    - to reproduce the paper, including nearly all figures and tables, type `make` or `make jasa2016.Rnw` at the command line
    - to reproduce Table 2, **Mike to fill in**
    - to reproduce the trace plots, run `Rscript 'R/trace_plots.R'`
    - to rerun the MCMC chains, run `make mcmc`
    - to redo the data wrangling, run `make bigfour`
    - to recompile the game results, run `make results`


## Notes

Please do not hesitate to contact us for help reproducing our results. 