---
title: "A unified approach to understanding randomness in sport"
output:
  github_document:
    html_preview: true
---
 
A paper analyzing competitiveness across the four major sports.

To knit the paper:

#. Use the **Compile PDF** button in RStudio, or
#. Select **Tools -> Shell...** from the menu and execute:
```{r, engine='bash', eval=FALSE}
make
```

## Data wrangling

To re-build the `bigfour` data:
```{r, engine='bash', eval=FALSE}
make bigfour
```

Note that this operation will look for a set of proprietary data files in a `data_raw` folder. This folder is not present in this repository due to licensing restrictions. However, our version of the resulting `bigfour.rda` file is present in the repository. This will enable you to run our simulations using the code below. 

To re-do the simulations:
```{r, engine='bash', eval=FALSE}
make mcmc
```

## Compile papers

To make the Sloan paper:
```{r, engine='bash', eval=FALSE}
# install dependencies for Ubuntu
# sudo apt get install texlive-xetex texlive-math-extra
make sloan2016.pdf
```
Note that this requires `xelatex` and the Cambria family of fonts installed. 

To make the journal paper:
```{r, engine='bash', eval=FALSE}
make aoas2017.pdf
```

